---
title: "ANA 515 Assignment 2"
author: "sunil salunke"
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
date: "6/19/2022"
output:
  html_document:
    theme:
      bootswatch: flatly
  word_document: default
  pdf_document: default
---

## Section 1: Description of the data

This data set contains information about airlines and their safety incidents. This data is measuring information about safety incidents, fatal accidents, and fatalities over duration of 1985 to 2014 across 56 airlines. This safety information is divided in 3 ways. Incidents column has information about number of safety incidents with the airline. Fatal accidents column has information about number of fatal accidents. Fatalities column has information about total number of fatalities. This safety information is divided into two time periods i.e. 1985 to 1999 and 2000 to 2014. This data set also has information about available seat km flown every week.

The data source for this information is from aviation-safety.net website. This website is maintained by flight safety foundation. Some of the questions that I would like to answer are, should travelers avoid flying airlines that has crash in the past? Is there a correlation and consistency between number of fatalities from one period to another? What are the number of fatalities by each accident? Is there a correlation and consistency between number of incidents from one period to another? 

This data is saved in CSV file format on github. This is a flat file. This is a delimited file and comma is the delimiter used in the file.

## Section 2: Reading the data into R
```{r echo=TRUE}
#using read.csv to read data from csv file from a URL, 
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/airline-safety/airline-safety.csv"
airline_safety_df <- read.csv(url)

#head of output data frame airline safety
knitr::kable(head(airline_safety_df))
```

#### comment:
I have used read.csv function to read the data from github url. This is a base r function with in-built functionality to read CSV file.


## Section 3: Clean the data

```{r message=FALSE, warning=FALSE}
#import packages
library(tidyverse)
library(summarytools)
```

```{r echo=TRUE}
#let's clean data by creating a separate data frame for 1985-1999
airline_safety_df_85_99 <- select(airline_safety_df,airline, avail_seat_km_per_week, incidents_85_99, fatal_accidents_85_99, fatalities_85_99)

#let's rename columns incidents_85_99 to incidents, fatal_accidents_85_99 to fatal_accidents, and fatalities_85_99 to fatalities in airline_safety_df_85_99 data frame
airline_safety_df_85_99 = airline_safety_df_85_99 %>% rename(incidents = incidents_85_99, fatal_accidents = fatal_accidents_85_99, fatalities = fatalities_85_99)

#let's check output data frame for 1985-1999 with columns renamed
knitr::kable(head(airline_safety_df_85_99))

#let's arrange our data by number of incidents in descending manner
airline_safety_df_85_99 <- airline_safety_df_85_99 %>% arrange(desc(incidents))

#let's check output data frame for 1985-1999 with incidents arranged in descending manner
knitr::kable(head(airline_safety_df_85_99))

#let's clean  data by creating a separate data frame for 2000-2014
airline_safety_df_00_14 <- select(airline_safety_df,airline, avail_seat_km_per_week, incidents_00_14, fatal_accidents_00_14, fatalities_00_14)

#let's rename columns incidents_00_14 to incidents, fatal_accidents_00_14 to fatal_accidents, and fatalities_00_14 to fatalities in airline_safety_df_00_14 data frame
airline_safety_df_00_14 = airline_safety_df_00_14 %>% rename(incidents = incidents_00_14, fatal_accidents = fatal_accidents_00_14, fatalities = fatalities_00_14)

#let's check output data frame for 1985-1999 with columns renamed
knitr::kable(head(airline_safety_df_00_14))
```

## Section 4: Characteristics of the data

This data frame has `r nrow(airline_safety_df)` rows and `r ncol(airline_safety_df)` columns. The names of the columns and a brief description of each are in the table below:

## A descriptive table:

```{r echo= TRUE} 
library(knitr)
columns_summary <- data.frame(
Columns = c(colnames(airline_safety_df)),
Description = c(
"Airline (asterisk indicates that regional subsidiaries are included)","Available seat kilometers flown every week", "Total number of incidents, 1985–1999", "Total number of fatal accidents, 1985–1999", "Total number of fatalities, 1985–1999", "Total number of incidents, 2000–2014", "Total number of fatal accidents, 2000–2014", "Total number of fatalities, 2000–2014")
)

kable(columns_summary, caption = "column_summary")
```

## Section 5: Summary statistics
## Subset the dataset:
## Picking three columns to use summary function:
```{r echo=TRUE}
airline_safety_df_subset <- select(airline_safety_df, incidents_85_99, fatal_accidents_85_99, fatalities_85_99)
```

## Produce a summary of the subset: checking for missing values
```{r echo=TRUE}
kable(airline_safety_df_subset %>% 
  summarise_all(list(missing_count = ~ sum(is.na(.)))))
```
##### There are no missing values in above 3 columns.

## Produce a summary of the subset:summary statistics
```{r, include = TRUE}
Summarytable<-summary(airline_safety_df_subset) #creates the summary
Summarytable #prints the summary in the output
```

##### Above is the summary of these 3 columns.

##### There are no missing values in above 3 columns.

##### For 1985–1999 period, across different airlines, total number of minimum incidents are 0 and total number of maximum incidents are 76. 

##### For 1985–1999 period, across different airlines, total number of minimum fatal accidents are 0 and total number of maximum fatal accidents are 14. 

##### For 1985–1999 period, across different airlines, total number of minimum fatalities are 0 and total number of maximum fatalities are 535. 

